---
title: "3. How to draw a pedigree"
author: "Sheng Luan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. How to draw a pedigree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 6.5,
  dpi = 300,
  out.width = "100%"
)
library(visPedigree)
```

1. [Drawing the pedigree graph](#1)   
1.1 [A simple pedigree graph](#1-1)   
1.1.1 [Highlighting specific individuals](#1-1-1)   
1.1.2 [Showing inbreeding coefficients](#1-1-2)   
1.2 [A reduced pedigree graph](#1-2)   
1.3 [An outlined pedigree graph](#1-3)   
1.4 [How to use this package in a selective breeding program](#1-4)   
1.4.1 [Analysis of founders for an individual](#1-4-1)   
1.4.2 [The contribution of different families in a selective breeding program](#1-4-2) 

## 1 Drawing the pedigree {#1}
The **visped** function takes a pedigree tidied by the `tidyped` function and outputs a hierarchical graph for all individuals in the pedigree. The graph can be displayed on the default graphics device and saved as a PDF file. The graph in the PDF file is a vector drawing, which is legible and avoids overlapping. It is especially useful when the number of individuals is large or when individual labels are long. This function can visualize very large pedigrees (> 10,000 individuals per generation) by compacting full-sib individuals. It is particularly effective for aquatic animal pedigrees, which typically include many full-sib families per generation in the nucleus breeding population. A pedigree outline without individual labels is shown if the graph width exceeds the maximum PDF width (200 inches). This helps breeders quickly review the population construction process and identify any introduction of new genetic material.

**Important Note:** It is strongly recommended to set the `cand` parameter when tidying a pedigree. Pruning the pedigree by specifying candidates allows for more accurate generation inference and a more logical layout in the resulting pedigree tree.

Additionally, isolated individuals (those with no parents and no progeny) are automatically filtered out by `visped()` to prevent cluttering the graph. These individuals are assigned Generation 0 during the tidying process.

A small pedigree is drawn in the following figure. The following code also demonstrates how to save the graph as a high-quality vector graphic in a PDF file.

```{r smallped, fig.show="hold"}
tidy_small_ped <-
  tidyped(ped = small_ped,
          cand = c("Y", "Z1", "Z2"))
visped(tidy_small_ped, compact = TRUE, file = tempfile(fileext = ".pdf"))
```

In the above graph, two shapes and three colors are used. Circles represent individuals, while squares represent families. Dark sky blue indicates males, dark goldenrod indicates females, and dark olive green indicates unknown sex. For example, a dark sky blue circle represents a male individual, while a dark goldenrod square represents all female individuals in a full-sib family when `compact = TRUE`. The ancestors are drawn at the top and descendants are drawn at the bottom in the pedigree graph. Parents and offspring are connected via dummy nodes. Lines from offspring to dummy nodes are dark grey, while lines from dummy nodes to parents match the parents' respective colors.

### 1.1 A simple pedigree graph {#1-1}
The trimmed **simple_ped** pedigree is drawn and displayed on the default graphics device. The **addgen** and **addnum** parameters need to be set to TRUE when tidying the pedigree using the **tidyped** function.
```{r vissimpleped, fig.show="hold"}
tidy_simple_ped <- tidyped(simple_ped)
visped(tidy_simple_ped, cex=0.3, symbolsize=10)
```

Generation labels can be added on the left margin of the pedigree graph by setting `genlab = TRUE`. This is useful for deep pedigrees when you want to quickly identify the generation of each row.

```{r vissimpleped_genlab, fig.show="hold"}
visped(tidy_simple_ped, cex = 0.3, symbolsize = 10, genlab = TRUE)
```

In deep pedigrees the generation labels may appear small because their size is tied to the node scaling. The `genlabcex` parameter lets you set the generation-label size independently of the individual-label size (`cex`). For example, when `cex` is deliberately small to fit many individuals, you can still keep the generation labels readable:

```{r vissimpleped_genlabcex, fig.show="hold"}
# cex controls individual label size; genlabcex controls generation label size
visped(tidy_simple_ped, cex = 0.3, symbolsize = 10, genlab = TRUE, genlabcex = 1.2)
```

Figures displayed in the RStudio Plots panel often have limited resolution. Individual IDs may overlap if the pedigree is large and the plot area is restricted. This can be resolved by saving the graph as a vector graphic in a PDF file. The **visped** function will suppress output to the default device if `showgraph = FALSE`.

```{r}
suppressMessages(visped(tidy_simple_ped, showgraph = FALSE, file = tempfile(fileext = ".pdf")))
```

By setting the **file** parameter, you can generate a high-definition PDF version of the pedigree.

#### 1.1.1 Highlighting specific individuals {#1-1-1}
Specific individuals can be highlighted in the pedigree graph using the **highlight** parameter. This is useful for marking candidates, founders, or any individuals of interest.

You can provide a character vector of individual IDs to use the default highlight colors (purple border and light purple fill):
```{r highlight1}
visped(tidyped(small_ped), highlight = c("Y", "Z1"))
```

You can also highlight an individual and its relatives (ancestors and descendants) by setting `trace = TRUE`:
```{r highlight_trace}
# Highlight individual "Y" and all its ancestors and descendants
visped(tidyped(small_ped), highlight = "Y", trace = TRUE)
```

Alternatively, you can customize the colors by providing a list with **ids**, **frame.color**, and **color**:
```{r highlight2}
visped(tidyped(small_ped), 
       highlight = list(ids = c("Y", "Z1"), 
                        frame.color = "#4caf50", 
                        color = "#81c784"))
```

### 1.1.2 Showing inbreeding coefficients {#1-1-2}
Inbreeding coefficients can be displayed on the pedigree graph using the **showf** parameter in the **visped** function. This requires that the pedigree has been processed with inbreeding coefficients calculated using the **inbreed** parameter in the **tidyped** function.
```{r showinbreed}
library(data.table)
test_ped <- data.table(
  Ind = c("A", "B", "C", "D", "E"),
  Sire = c(NA, NA, "A", "C", "C"),
  Dam = c(NA, NA, "B", "B", "D"),
  Sex = c("male", "female", "male", "female", "male")
)
tidy_test_ped_inbreed <- tidyped(test_ped, inbreed = TRUE)
visped(tidy_test_ped_inbreed, showf = TRUE)
```

### 1.2 A reduced pedigree graph {#1-2}

Warning messages will be shown when you try to draw the pedigree graph of the deep_ped dataset.
```{r deepped, eval=FALSE}
cand_J11_labels <- deep_ped[(substr(Ind, 1, 3) == "K11"), Ind]
visped(
  tidyped(deep_ped,
    cand = cand_J11_labels,
    tracegen = 3),
  cex=0.08,
  symbolsize=5.5
  )
```

```
  Too many individuals (>=3362) in one generation!!! Two choices:
1. Removing full-sib individuals using the parameter compact = TRUE; or, 
2. Visualizing all nodes without labels using the parameter outline = TRUE.
Rerun visped() function!
```

The function indicates that there are too many individuals in a single generation to draw a standard pedigree graph. It is recommended to use the **compact** or **outline** parameters to simplify the pedigree.

First, let's try the **compact** parameter and output it to a PDF file. The plot on the default device may suffer from significant overlapping due to the high density of individuals.
```{r reduceped1}
cand_J11_labels <- deep_ped[(substr(Ind,1,3) == "K11"),Ind]
visped(
  tidyped(
    deep_ped,
    cand = cand_J11_labels,
    trace = "up",
    tracegen = 3
  ),
  cex=0.08, 
  symbolsize=5.5,
  compact = TRUE,
  showgraph = TRUE,
  file = tempfile(fileext = ".pdf")
)
```

You can open the generated PDF file to view the high-definition pedigree vectorgraph. Most of shapes are square at bottom, and the internal numbers are the total number of male or female individuals for each family. Individual labels may be shorter than the shapes and might not align perfectly. Individual labels can be resized using the `cex` parameter. The `cex` parameter controls the font size of individual IDs. Increasing `cex` makes the labels larger, while decreasing it makes them smaller. The `cex` value typically ranges from 0 to 1, but can be larger; adjustments of 0.1 are usually sufficient. The **visped** function will output warning messages including the cex value which was used for drawing the pedigreed graph. 

```{r reduceped2}
visped(
  tidyped(
    deep_ped,
    cand = cand_J11_labels,
    trace = "up",
    tracegen = 3
  ),
  compact = TRUE,
  cex = 0.83,
  showgraph = FALSE,
  file = tempfile(fileext = ".pdf")
)
```
You can open the generated PDF file to view the high-definition pedigree vectorgraph. The labels align better with the shapes compared to the previous version. You can continue to adjust `cex` until the labels are sized appropriately.

### 1.3 An outlined pedigree graph {#1-3}
Setting `outline = TRUE` produces an outlined pedigree graph. Individual labels will not be shown in the graph. This is highly effective for large pedigrees with many individuals.

The following code generates an outlined pedigree graph in a PDF file.

```{r reduceped3}
suppressMessages(visped(
  tidyped(
    deep_ped,
    cand = cand_J11_labels, 
    tracegen = 3),
  compact = TRUE,
  outline = TRUE,
  showgraph = TRUE,
  file = tempfile(fileext = ".pdf")
))
```

### 1.4 How to use this package in a selective breeding program {#1-4}    
#### 1.4.1 An analysis of founders for an individual {#1-4-1}    
Selective breeding is a process of enriching desirable minor genes from multiple founders through successive generations of mating. This is supported by the well-known infinitesimal model (or minor polygene hypothesis).

We select the individual "K110550H" in the deep_ped dataset to visualize its pedigree. The following code generates the pedigree graph for a specific individual in a PDF file.
```{r pedofoneind}
suppressWarnings(
  K110550H_ped <- tidyped(deep_ped, cand = "K110550H")
)
if(interactive()) {
  visped(K110550H_ped, compact = TRUE)
}
```

As you can see from the figure above, the number of founder individuals (without parents) of the K110550H individual is `r nrow(K110550H_ped[is.na(Sire) & is.na(Dam)])`. This indicates that the individual has accumulated favorable genes from many founders, contributing to genetic gain in the target traits.

#### 1.4.2 The contribution of different families in a selective breeding program {#1-4-2}

Under optimum contribution theory, families contribute different numbers of individuals to the next generation, with higher-indexing families contributing more. By visualizing pedigree, we can directly see the contribution ratio of different families.

The code below shows the parental composition of 106 families born in the nucleus breeding population in 2007. Setting `tracegen = 2` limits the graph to two generations (parents and grandparents).
```{r optiMate}
cand_2007_G8_labels <-
  big_family_size_ped[(Year == 2007) & (substr(Ind, 1, 2) == "G8"), Ind]
suppressWarnings(
  cand_2007_G8_tidy_ped_ancestor_2 <-
    tidyped(
      big_family_size_ped,
      cand = cand_2007_G8_labels,
      trace = "up",
      tracegen = 2
    )
)
sire_label <-
  unique(cand_2007_G8_tidy_ped_ancestor_2[Ind %in% cand_2007_G8_labels,
                                          Sire])
dam_label <-
  unique(cand_2007_G8_tidy_ped_ancestor_2[Ind %in% cand_2007_G8_labels,
                                          Dam])
sire_dam_label <- unique(c(sire_label, dam_label))
sire_dam_label <- sire_dam_label[!is.na(sire_dam_label)]
sire_dam_ped <-
  cand_2007_G8_tidy_ped_ancestor_2[Ind %in% sire_dam_label]
sire_dam_ped <-
  sire_dam_ped[, FamilyID := paste(Sire, Dam, sep = "")]
family_size <- sire_dam_ped[, .N, by = c("FamilyID")]
fullsib_family_label <- unique(sire_dam_ped$FamilyID)
suppressMessages(
  visped(
    cand_2007_G8_tidy_ped_ancestor_2,
    compact = TRUE,
    outline = TRUE,
    showgraph = TRUE
  )
)
```

In the above figure, 106 families are shown at bottom, the parents are shown in middle, and the grandparents are shown at top. It can be seen that the parents are composed of `r length(sire_label[!is.na(sire_label)])` sires and `r length(dam_label[!is.na(dam_label)])` dams. The parents are from `r length(fullsib_family_label)` full-sib families in the generation of grandparent. Approximately `r family_size$N[1]+family_size$N[2]` parents originate from just two full-sib families due to the application of optimum contribution theory, accounting for `r round((family_size$N[1]+family_size$N[2])/sum(family_size$N),4)*100`% of all parents.