---
title: "2. Efficient visPedigree Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Efficient visPedigree Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 6,
  dpi = 96,
  out.width = "100%"
)
```

This vignette summarizes efficient day-to-day workflows for `visPedigree`
after the `tidyped` architecture updates. The goal is simple:

1. tidy once,
2. reuse the resulting `tidyped` object many times,
3. subset safely,
4. trace candidates explicitly when pedigree completeness matters.

For basic tidying, see `tidy-pedigree`. For downstream statistics, see
`pedigree-analysis`.

## 1. Load packages and example data

```{r setup}
library(visPedigree)
library(data.table)

data(simple_ped, package = "visPedigree")
```

## 2. Tidy once, reuse many times

The most efficient workflow is to create a master `tidyped` object once and
reuse it for plotting, tracing, inbreeding, and matrix calculations.

```{r tidy-once}
tp_master <- tidyped(simple_ped)

class(tp_master)
is_tidyped(tp_master)
pedmeta(tp_master)
```

This avoids repeated validation, founder insertion, loop checking, generation
assignment, and integer re-indexing.

## 3. Fast repeated tracing from an existing `tidyped`

When the input is already a `tidyped` object and `cand` is supplied,
`tidyped()` now uses a fast path. It skips the expensive global preprocessing
steps and directly traces the requested candidates.

```{r fast-trace}
tp_up <- tidyped(tp_master, cand = "J5X804", trace = "up", tracegen = 2)
tp_down <- tidyped(tp_master, cand = "J0Z990", trace = "down")

has_candidates(tp_up)
tp_up[, .(Ind, Sire, Dam, Cand)]
```

Recommended pattern:

```{r fast-trace-pattern, eval = FALSE}
# expensive once
# tp_master <- tidyped(raw_ped)

# cheap many times
# tp_a <- tidyped(tp_master, cand = ids_a, trace = "up")
# tp_b <- tidyped(tp_master, cand = ids_b, trace = "all", tracegen = 3)
# tp_c <- tidyped(tp_master, cand = ids_c, trace = "down")
```

## 4. Safe `data.table` usage on `tidyped`

A `tidyped` object is also a `data.table`, so by-reference workflows remain
available.

### 4.1 Adding new columns is safe

```{r dt-modify}
tp_work <- copy(tp_master)
tp_work[, phenotype := seq_len(.N)]

class(tp_work)
head(tp_work[, .(Ind, phenotype)])
```

The `tidyped` class is preserved after `:=` operations.

### 4.2 Incomplete row subsetting now degrades safely

If row filtering removes required parents, the result is no longer a complete
pedigree. In that case the object is downgraded to a plain `data.table` with a
warning.

```{r incomplete-subset}
ped_year <- data.table(
  Ind = c("A", "B", "C", "D"),
  Sire = c(NA, NA, "A", "C"),
  Dam = c(NA, NA, "B", "B"),
  Year = c(2000, 2000, 2005, 2006)
)

tp_year <- tidyped(ped_year)
sub_dt <- tp_year[Year > 2005]

class(sub_dt)
sub_dt
```

This behavior prevents invalid integer pedigree indices from silently reaching
C++ code.

### 4.3 Use explicit tracing when you need a valid sub-pedigree

If the goal is to keep a structurally valid pedigree around focal individuals,
use candidate tracing instead of ad hoc row filtering.

```{r explicit-tracing}
valid_sub_tp <- tidyped(tp_year, cand = "D", trace = "up")

class(valid_sub_tp)
valid_sub_tp[, .(Ind, Sire, Dam, Cand)]
```

## 5. `splitped()` versus `pedsubpop()`

These two functions serve different purposes.

- `splitped()` returns the actual split pedigree objects.
- `pedsubpop()` returns a summary table.

```{r split-vs-summary}
sub_tps <- splitped(tp_master)
length(sub_tps)
class(sub_tps[[1]])

pedsubpop(tp_master)
```

Use `splitped()` when you need downstream analysis on each component. Use
`pedsubpop()` when you only need the component summary.

## 6. Use accessors instead of manual attribute checks

The updated accessors are the preferred way to inspect object state.

```{r accessors}
tp_f <- inbreed(tp_master)

is_tidyped(tp_f)
has_inbreeding(tp_f)
has_candidates(tp_f)
pedmeta(tp_f)
```

This is preferable to hand-written checks such as `"f" %in% names(tp)` or
manual attribute access scattered throughout user code.

## 7. Recommended high-efficiency workflow

A practical pattern for large pedigrees is:

```{r best-practice, eval = FALSE}
# 1. build one validated master object
# tp_master <- tidyped(raw_ped)

# 2. add analysis-specific columns in place
# tp_master[, phenotype := pheno_vector]
# tp_master[, cohort := year_vector]

# 3. extract valid candidate sub-pedigrees explicitly
# tp_sel <- tidyped(tp_master, cand = selected_ids, trace = "up", tracegen = 3)

# 4. run downstream analysis on either the full master or traced sub-pedigree
# pedstats(tp_master)
# pedmat(tp_sel)
# inbreed(tp_sel)
# visped(tp_sel)

# 5. split only when disconnected components really matter
# comps <- splitped(tp_master)
```

## 8. Practical rules of thumb

1. Call `tidyped()` on raw pedigree data once.
2. Reuse the resulting `tidyped` object as the master pedigree.
3. Use `tidyped(tp_master, cand = ...)` for valid local extraction.
4. Use ordinary row filtering only when a plain `data.table` result is acceptable.
5. Use `splitped()` for actual component objects and `pedsubpop()` for summaries.
6. Use `pedmeta()`, `is_tidyped()`, `has_inbreeding()`, and `has_candidates()` to inspect object state.

These rules keep workflows fast, explicit, and structurally safe.
