---
title: "Pedigree Analysis and Population Genetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pedigree Analysis and Population Genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a comprehensive workflow for analyzing population genetics using the `visPedigree` package. It introduces the unified `pediv()` framework, tracks the evolution of inbreeding, measures genetic diversity, and estimates effective population sizes based on robust mathematical models.

## 1. Setup and Data Preparation

We begin by loading a simulated historical pedigree.

```{r setup}
library(visPedigree)
library(data.table)

# Use a simulated complex pedigree provided with the package
data(deep_ped, package = "visPedigree")

# Tidy the pedigree: This automatically computes generations and sorts appropriately
tped <- tidyped(ped = deep_ped)
```

## 2. Demographic Overview and Generation Intervals

Good population management starts with understanding demographic turnover. `pedstats()` returns the structural depth and census sizes. The expected equivalent complete generations (ECG, Maignel et al., 1996) provides a measure of pedigree completeness, defined as:

$$ ECG = \sum_{i=1}^n \left(\frac{1}{2}\right)^{g_i} $$
where $n$ is the number of ancestors and $g_i$ is the number of generations separating the individual from its ancestor.

```{r stats}
# General Statistics
stats <- pedstats(tped)
print(stats)
```

*(Note: Provide a `timevar` such as `BirthYear` to compute generation intervals, providing insight into the age of parents when offspring are born).*

## 3. Comprehensive Diversity Metrics using `pediv`

Instead of running separate components, `visPedigree` offers the `pediv()` wrapper function to conduct a sweeping analysis of a target reference population. The metrics include genetic contributions ($f_a$, $f_e$) and Effective Population Sizes ($N_e$).

Let's define a **reference population** (e.g., individuals in the last two generations) to evaluate.

```{r define_ref}
# Identify our cohort of interest (e.g., modern descendants)
ref_pop <- tped[Gen >= max(Gen) - 1, Ind]
length(ref_pop)
```

```{r pediv_analysis}
# Run comprehensive population diversity analysis
div_res <- pediv(tped, reference = ref_pop, top = 10, ncores = 1, seed = 42L)

# Summary table provides an elegant overview
print(div_res$summary)
```

### 3.1 Founder Equivalents ($f_g$, $f_a$ and $f_e$)

The metrics $f_g$ (Caballero & Toro, 2000), $f_a$ (Boichard et al., 1997), and $f_e$ (Lacy, 1989) evaluate how unbalanced the founding genome has become:

- **Founder Genome Equivalents ($f_g$)**:
  $$ f_g = \frac{1}{2\bar{C}} $$
  where $\bar{C}$ is the mean coancestry of the reference population. $f_g$ accounts for unequal founder contributions, bottlenecks, and the loss of alleles due to random genetic drift. It strictly follows the hierarchy $f_g \leq f_a \leq f_e \leq N_{founders}$.
  
- **Effective Number of Founders ($f_e$)**: 
  $$ f_e = \frac{1}{\sum_{i=1}^f p_i^2} $$
  where $p_i$ is the expected genetic contribution of founder $i$. It assumes all founders contribute independently.
  
- **Effective Number of Ancestors ($f_a$)**: 
  $$ f_a = \frac{1}{\sum_{j=1}^a q_j^2} $$
  where $q_j$ is the **marginal** genetic contribution of ancestor $j$ not explained by previously selected ancestors. This metric directly addresses population bottlenecks.

The top ancestors driving these metrics can be investigated:
```{r top_ancestors}
print(div_res$ancestors)
```

### 3.2 Effective Population Size ($N_e$)

$N_e$ reflects the size of an idealized population that would experience the same rate of genetic drift or inbreeding as the observed population. `pediv()` calculates three variations:

1. **Demographic $N_e$**: Based purely on the census of breeding males and females.
2. **Inbreeding $N_e$**: Driven by the individual rate of inbreeding ($\Delta F_i$).
3. **Coancestry $N_e$**: Computed from the rate of coancestry ($\Delta c$) between pairs in the reference group, generally yielding the most strictly conservative estimate in heavily managed setups.

## 4. Inbreeding Evolution ($F$)

If we wish to extract granular, individual-level metrics, we can calculate Wright's (1922) coefficient of inbreeding ($F$).

$$ F_{XY} = \sum \left( \frac{1}{2} \right)^{n_1 + n_2 + 1} (1 + F_A) $$

```{r inbreed_analysis}
# Calculate Individual F
tped_f <- inbreed(tped)

# Evaluate average inbreeding by generation
f_trend <- tped_f[, .(MeanF = mean(f, na.rm=TRUE)), by = Gen]
print(f_trend)
```

Using `pedinbreed_class()`, we can stratify the reference population by their inbreeding severity.

```{r F_class}
f_class <- pedinbreed_class(tped_f)
print(f_class)
```

## 5. Advanced: Partial Inbreeding and Gene Flow

### 5.1 Gene Flow (Ancestry Proportions)
Track the segregation of traits across historical founder groups. E.g., if founders are tagged by geographic origin:

```{r gene_flow}
# Simulate 'origin' labels for founders (Gen = 1)
tped_dated <- copy(tped)
founders <- tped_dated[Gen == 1, Ind]
tped_dated[Ind %in% founders[1:5], Origin := "LineA"]
tped_dated[Ind %in% founders[6:10], Origin := "LineB"]

# Gene flow traces LineA and LineB presence in descendants
ancestry <- pedancestry(tped_dated, labelvar = "Origin")
print(head(ancestry))
```

### 5.2 Partial Inbreeding ($F_{ij}$)
The total inbreeding coefficient $F_i$ can be partitioned to isolate the blame of specific shared ancestors (Gulisija & Crow, 2007).

```{r partial_f}
# Select highly inbred descendants to query
target_inbred <- tped_f[f > 0.1, Ind][1]

# Decompose inbreeding mapped to the top historical ancestors
partial_f <- pedpartial(tped_f, ancestors = div_res$ancestors$Ind, top = 5)

# View exactly which ancestors caused F in our target
print(partial_f[Ind == target_inbred])
```

## Conclusion

By adopting `pediv()`, demographic summaries, and partial tracing, breeders can pinpoint bottleneck risks, mitigate rapid coancestry accumulation, and curate genetically resilient populations.