---
title: "Pedigree Analysis Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pedigree Analysis Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a complete workflow for pedigree analysis using the `visPedigree` package. We will cover basic statistics, inbreeding monitoring, population structure analysis, and advanced ancestral tracking using the implemented gene flow and partial inbreeding algorithms.

## 1. Setup and Data Preparation

First, we load the package and prepare a sample pedigree. We will use a simulated pedigree `simple_ped` included in the package, but for this demonstration, we'll simulate a slightly more complex structure with inbreeding events to showcase the features.

```{r setup}
library(visPedigree)
library(data.table)

# Simulate a pedigree with 5 generations
# Foundation: 2 males (S1, S2), 2 females (D1, D2)
ped_data <- data.table(
  Ind = c("S1", "S2", "D1", "D2", 
          "G1_1", "G1_2", "G1_3", "G1_4",
          "G2_1", "G_Inbred", 
          "G3_1"),
  Sire = c(NA, NA, NA, NA, 
           "S1", "S1", "S2", "S2", 
           "G1_1", "G1_1", 
           "G2_1"),
  Dam = c(NA, NA, NA, NA, 
          "D1", "D2", "D1", "D2", 
          "G1_3", "G1_3", # G_Inbred is G1_1 x G1_3 (Full Sibs if S1,D1? No. G1_1=S1xD1, G1_3=S2xD1. Half-sibs)
          "G_Inbred"),
  Sex = c("male", "male", "female", "female",
          "male", "male", "female", "female",
          "male", "female",
          "male"),
  BirthYear = c(2010, 2010, 2010, 2010,
                2012, 2012, 2013, 2013,
                2015, 2016,
                2019)
)

# Tidy the pedigree (sorts, adds generations, checks integrity)
tped <- tidyped(ped_data)
print(tped)
```

## 2. Basic Statistics & Generation Intervals

Before deeper analysis, we look at the general structure and reproductive cycle speed.

```{r stats}
# General Statistics
stats <- pedstats(tped, timevar = "BirthYear", unit = "year")
print(stats)

# The 'stats' object contains:
# - Summary counts (N, Sires, Dams)
# - Ancestral depth (ECG) summary
# - Generation intervals (if birth dates provided)
```

## 3. Inbreeding Analysis

Monitoring inbreeding is crucial for population management.

### Calculate Inbreeding Coefficients (F)
```{r inbreed}
# Calculate F
tped <- inbreed(tped)

# Show individuals with non-zero inbreeding
print(tped[f > 0, .(Ind, Sire, Dam, Gen, f)])
```

### Stratification
Classify the population by inbreeding levels.
```{r inbreed_class}
# Summary of inbreeding levels
f_class <- pedinbreed_class(tped, method = "summary")
print(f_class)
```

### Effective Population Size (Ne)
Estimate the effective population size ($N_e$), which is a key metric for monitoring genetic diversity. `visPedigree` provides three methods: based on coancestry (`"coancestry"`, the default), rate of inbreeding (`"inbreeding"`), or demographic counts (`"demographic"`).

You can calculate $N_e$ grouped by year or generation:
```{r ne_by_year}
# Estimate Ne per cohort (grouped by BirthYear)
ne_est <- pedne(tped, by = "BirthYear", method = "inbreeding")
print(ne_est)
```

Alternatively, you can evaluate the overall $N_e$ for a specific subset (e.g., the last generation) by omitting the `by` parameter:
```{r ne_overall}
# Get the IDs of the latest generation
latest_gen_ids <- tped[Gen == max(Gen), Ind]

# Calculate overall Ne for this specific candidate group using the default coancestry method
overall_ne <- pedne(tped, cand = latest_gen_ids)
print(overall_ne)
```

## 4. Genetic Diversity & Relationships

Track the average relationship within the population. Increasing mean relationship indicates a loss of genetic diversity.

```{r rel}
# Mean relationship by generation
avg_rel <- pedrel(tped, by = "Gen")
print(avg_rel)
```

## 5. Ancestral Analysis

Identify which ancestors are most influential and trace gene flow.

### Most Influential Ancestors
Find ancestors with the highest marginal genetic contribution.
```{r contrib}
# Calculate contributions to the last generation
last_gen_ids <- tped[Gen == max(Gen), Ind]
contributions <- pedcontrib(tped, cohort = last_gen_ids, mode = "ancestor")

print(contributions)
```

### Gene Flow / Ancestry Proportions
If our founders came from different origins (e.g., Line A vs Line B), we can trace their proportions.

```{r gene_flow}
# Assign arbitrary labels to founders for demonstration
tped_dated <- copy(tped)
tped_dated[Ind %in% c("S1", "D1"), Origin := "LineA"]
tped_dated[Ind %in% c("S2", "D2"), Origin := "LineB"]

# Calculate proportion of LineA vs LineB in all individuals
ancestry <- pedancestry(tped_dated, labelvar = "Origin")
print(tail(ancestry))
```

## 6. Advanced: Partial Inbreeding

Partial inbreeding ($F_{ij}$) allows us to decompose an individual's total inbreeding coefficient into the specific contributions from each ancestor. This answers: *"Who is responsible for the inbreeding of this animal?"*

```{r partial}
# Let's analyze the inbreeding of the last individual (G3_1)
target_ind <- "G3_1"
total_f <- tped[Ind == target_ind, f]
cat("Total Inbreeding for", target_ind, ":", total_f, "\n")

# Calculate partial inbreeding from top ancestors
partial_f <- pedpartial(tped, ancestors = c("S1", "D1", "S2", "D2"), top = 5)

# Extract result for the target individual
res <- partial_f[Ind == target_ind]
print(res)
```

## Summary

`visPedigree` provides a full suite of tools for:
1.  **Demography**: Stats and intervals.
2.  **Health**: Inbreeding and diversity monitoring.
3.  **Structure**: Ancestral contributions and gene flow.
4.  **Visualization**: (See `vignette("draw-pedigree")` for plotting details).

These tools enable efficient management of breeding programs and conservation genetics.
