---
title: "Pedigree Analysis Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pedigree Analysis Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a complete workflow for pedigree analysis using the `visPedigree` package. We will cover basic statistics, inbreeding monitoring, population structure analysis, and advanced ancestral tracking using the implemented gene flow and partial inbreeding algorithms.

## 1. Setup and Data Preparation

First, we load the package and prepare a sample pedigree. We will use a simulated pedigree `simple_ped` included in the package, but for this demonstration, we'll simulate a slightly more complex structure with inbreeding events to showcase the features.

```{r setup}
# Ensure the development version is loaded instead of the globally installed one
devtools::load_all("../")
library(visPedigree)
library(data.table)

# Simulate a pedigree with 4 generations
# Foundation: 2 males (S1, S2), 2 females (D1, D2)
ped_data <- data.table(
  Ind = c("S1", "S2", "D1", "D2", 
          "G1_1", "G1_2", "G1_3", "G1_4",
          "G2_1", "G2_Inbred", "G2_2", 
          "G3_1", "G3_2", "G3_3"),
  Sire = c(NA, NA, NA, NA, 
           "S1", "S1", "S2", "S2", 
           "G1_1", "G1_1", "G1_2",
           "G2_1", "G2_1", "G2_1"),
  Dam = c(NA, NA, NA, NA, 
          "D1", "D2", "D1", "D2", 
          "G1_3", "G1_3", "G1_4",
          "G2_Inbred", "G2_Inbred", "G2_2"),
  Sex = c("male", "male", "female", "female",
          "male", "male", "female", "female",
          "male", "female", "female",
          "male", "female", "male"),
  BirthYear = c(2010, 2010, 2010, 2010,
                2012, 2012, 2013, 2013,
                2015, 2016, 2016,
                2019, 2019, 2019)
)

# Tidy the pedigree (sorts, adds generations, checks integrity)
tped <- tidyped(ped_data)
print(tped)
```

## 2. Basic Statistics & Generation Intervals

Before deeper analysis, we look at the general structure and reproductive cycle speed.

```{r stats}
# General Statistics
stats <- pedstats(tped, timevar = "BirthYear", unit = "year")
print(stats)

# The 'stats' object contains:
# - Summary counts (N, Sires, Dams)
# - Ancestral depth (ECG) summary
# - Generation intervals (if birth dates provided)
```

## 3. Inbreeding Analysis

Monitoring inbreeding is crucial for population management.

### Calculate Inbreeding Coefficients (F)
```{r inbreed}
# Calculate F
tped <- inbreed(tped)

# Show individuals with non-zero inbreeding
print(tped[f > 0, .(Ind, Sire, Dam, Gen, f)])
```

### Stratification
Classify the population by inbreeding levels.
```{r inbreed_class}
# Summary of inbreeding levels
f_class <- pedinbreed_class(tped)
print(f_class)
```

### Effective Population Size (Ne)
Estimate the effective population size ($N_e$), which is a key metric for monitoring genetic diversity. `visPedigree` provides three calculation methods to analyze $N_e$ from different population genetics perspectives.

We can analyze $N_e$ for individuals in later generations (e.g., $Gen \ge 3$) to compare these three methods side by side. We use `by = "Gen"` to group the outputs.

```{r ne_methods_compare}
# Identify candidates in later generations and assign a grouping variable
ne_candidates <- tped[Gen >= 3, .(Ind, Gen)]

# 1. Coancestry-based Ne (Default)
# Relies on the rate of coancestry ($\Delta c$) between pairs of individuals.
# Very robust for managed populations, sensitive to shared bloodlines.
ne_co <- pedne(tped, cand = ne_candidates$Ind, by = "Gen", method = "coancestry")
print(ne_co)

# 2. Inbreeding-based Ne
# Uses the individual rate of inbreeding ($\Delta F$). Evaluates how rapidly 
# individuals become inbred compared to their ancestors.
ne_in <- pedne(tped, cand = ne_candidates$Ind, by = "Gen", method = "inbreeding")
print(ne_in)

# 3. Demographic-based Ne
# Counts distinct sires and dams contributing to the cohort.
# Simplistic baseline, ignores complex historical relationships.
ne_de <- pedne(tped, cand = ne_candidates$Ind, by = "Gen", method = "demographic")
print(ne_de)
```

**Understanding the Differences:**
- **Demographic** fluctuates simply based on how many parent IDs are directly listed as sires and dams for the cohort.
- **Inbreeding** evaluates individual paths to shared ancestors. For deliberate crossing schemes, it tends to yield higher, more optimistic estimates ($\sim 4.01$).
- **Coancestry** accurately detects the true loss of genetic variance by checking relationships between *pairs* in the generation. It recognizes widespread shared grandparents and great-grandparents, appropriately yielding a more conservative and arguably more relevant estimate for modern populations ($\sim 2.90$).

## 4. Genetic Diversity & Relationships

Track the average relationship within the population. Increasing mean relationship indicates a loss of genetic diversity.

```{r rel}
# Mean relationship by generation
avg_rel <- pedrel(tped, by = "Gen")
print(avg_rel)
```

## 5. Ancestral Analysis

Identify which ancestors are most influential and trace gene flow. The algorithm estimates both the effective number of founders ($f_e$) and the effective number of ancestors ($f_a$). 

- $f_e$: Quantifies genetic variation assuming founders contribute equally and independently to the reference cohort.
- $f_a$: Accounts for bottlenecks in the pedigree, providing a more stringent measure of the genetic base by identifying the marginal contributions of specific influential ancestors.

### Most Influential Ancestors
Find ancestors with the highest marginal genetic contribution. We'll use a subset of a more complex pedigree to demonstrate `pedcontrib`.

```{r contrib}
# Provide candidates from the last generation to evaluate their genetic origins
last_gen_ids <- tped[Gen == max(Gen), Ind]

# mode = "both" calculates both founder and ancestor contributions
contributions <- pedcontrib(tped, cand = last_gen_ids, mode = "both", top = 5)

# The summary shows the cohort size, expected and actual founder/ancestor metrics
print(contributions$summary)

# The top ancestors identified by Boichard's algorithm
print(contributions$ancestors)
```

### Gene Flow / Ancestry Proportions
If our founders came from different origins (e.g., Line A vs Line B), we can trace their proportions.

```{r gene_flow}
# Assign arbitrary labels to founders for demonstration
tped_dated <- copy(tped)
tped_dated[Ind %in% c("S1", "D1"), Origin := "LineA"]
tped_dated[Ind %in% c("S2", "D2"), Origin := "LineB"]

# Calculate proportion of LineA vs LineB in all individuals
ancestry <- pedancestry(tped_dated, labelvar = "Origin")
print(tail(ancestry))
```

## 6. Advanced: Partial Inbreeding

Partial inbreeding ($F_{ij}$) allows us to decompose an individual's total inbreeding coefficient into the specific contributions from each ancestor. This answers: *"Who is responsible for the inbreeding of this animal?"*

```{r partial}
# Let's analyze the inbreeding of the last individual (G3_1)
target_ind <- "G3_1"
total_f <- tped[Ind == target_ind, f]
cat("Total Inbreeding for", target_ind, ":", total_f, "\n")

# Calculate partial inbreeding from top ancestors
partial_f <- pedpartial(tped, ancestors = c("S1", "D1", "S2", "D2"), top = 5)

# Extract result for the target individual
res <- partial_f[Ind == target_ind]
print(res)
```

## Summary

`visPedigree` provides a full suite of tools for:
1.  **Demography**: Stats and intervals.
2.  **Health**: Inbreeding and diversity monitoring.
3.  **Structure**: Ancestral contributions and gene flow.
4.  **Visualization**: (See `vignette("draw-pedigree")` for plotting details).

These tools enable efficient management of breeding programs and conservation genetics.
