---
title: "4. Pedigree Analysis and Population Genetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Pedigree Analysis and Population Genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette summarizes a practical workflow for pedigree analysis with
`visPedigree`, with emphasis on the interpretation of the main indicators used
in breeding and conservation genetics.

The discussion is organized around five questions:

1. How complete and deep is the pedigree?
2. How long is the generation cycle?
3. Has genetic diversity been eroded by unequal founder use, bottlenecks, or drift?
4. How large is the effective population size under different definitions?
5. Are relationship, inbreeding, subpopulation structure, and gene flow under control?

## 1. Setup and Data Preparation

Different package datasets are used for different analytical tasks.

- `deep_ped` is useful for pedigree depth and diversity summaries.
- `big_family_size_ped` contains a `Year` column and is suitable for generation intervals.
- `small_ped` is convenient for relationship examples.
- `inbred_ped` is convenient for inbreeding and classification examples.

```{r setup}
library(visPedigree)
library(data.table)

data(deep_ped, package = "visPedigree")
data(big_family_size_ped, package = "visPedigree")
data(small_ped, package = "visPedigree")
data(inbred_ped, package = "visPedigree")

tp_deep <- tidyped(deep_ped)
tp_small <- tidyped(small_ped)
tp_inbred <- tidyped(inbred_ped)
```

## 2. Pedigree Overview with `pedstats()`

`pedstats()` provides a compact structural summary with three components:

- `$summary`: pedigree size and parental structure.
- `$ecg`: pedigree completeness and ancestral depth.
- `$gen_intervals`: generation intervals, only when a usable time column is available.

Here `deep_ped` has no explicit birth-date column. Accordingly,
`pedstats()` returns `$summary` and `$ecg`, whereas `$gen_intervals` remains
`NULL`.

```{r pedstats-overview}
stats_deep <- pedstats(tp_deep)

stats_deep$summary
tail(stats_deep$ecg)
stats_deep$gen_intervals
```

Pedigree depth and pedigree time are related but distinct quantities. The former
is addressed by `pedecg()`, whereas the latter is addressed by `pedgenint()`.

## 3. Pedigree Completeness with `pedecg()`

Equivalent complete generations (ECG) summarize the amount of ancestral
information available for each individual:

$$
ECG_i = \sum_j \left(\frac{1}{2}\right)^{g_{ij}}
$$

where $g_{ij}$ is the number of generations between individual $i$ and ancestor
$j$. ECG increases with both pedigree depth and pedigree completeness.

In practice:

- `ECG` combines depth and completeness.
- `FullGen` counts how many fully known ancestral generations exist.
- `MaxGen` records the deepest known ancestral path.

`ECG` is especially useful because it provides the depth adjustment required by
realized effective population size estimators based on inbreeding or coancestry.

```{r pedecg-example}
ecg_deep <- pedecg(tp_deep)

ecg_deep[order(-ECG)][1:10]
```

## 4. Generation Intervals with `pedgenint()`

Generation interval is the age of a parent at the birth of its offspring:

$$
L = t_{offspring} - t_{parent}
$$

`pedgenint()` estimates this quantity for seven pathway summaries:

- `SS`, `SD`, `DS`, `DD`: sex-specific gametic pathways.
- `SO`, `DO`: sex-independent sire-offspring and dam-offspring pathways.
- `Average`: all parent-offspring pairs combined.

The function accepts `Date`/`POSIXct` columns, date strings, or numeric years.
When only an integer year is available, it is converted internally to
`YYYY-07-01` as a mid-year approximation.

```{r pedgenint-basic}
tp_time <- tidyped(big_family_size_ped)

genint_year <- pedgenint(tp_time, timevar = "Year", unit = "year")
genint_year
```

The optional `cycle` parameter adds `GenEquiv`, which compares the observed mean
interval with a target breeding cycle:

$$
GenEquiv_i = \frac{\bar{L}_i}{L_{cycle}}
$$

Values larger than 1 indicate that the realized generation interval exceeds the
target cycle.

```{r pedgenint-cycle}
genint_cycle <- pedgenint(tp_time, timevar = "Year", unit = "year", cycle = 1.2)

genint_cycle[Pathway %in% c("SS", "SD", "DS", "DD", "Average")]
```

## 5. Subpopulation Structure with `pedsubpop()`

Before interpreting diversity or relationship metrics, it is useful to check
whether the pedigree forms a single connected population or a mixture of
separate components.

`pedsubpop()` has two modes:

- `by = NULL`: summarize disconnected pedigree components.
- `by = "..."`: summarize the pedigree by an existing grouping variable.

```{r pedsubpop-example}
ped_demo <- data.table(
  Ind = c("A", "B", "C", "D", "E", "F", "G", "H"),
  Sire = c(NA, NA, "A", NA, NA, "E", NA, NA),
  Dam  = c(NA, NA, "B", NA, NA, NA, NA, NA),
  Sex = c("male", "female", "male", "female", "male", "female", "male", "female"),
  Batch = c("L1", "L1", "L1", "L1", "L2", "L2", "L3", "L3")
)

tp_demo <- tidyped(ped_demo)

pedsubpop(tp_demo)
pedsubpop(tp_demo, by = "Batch")
```

This is a compact summary tool. When the actual split pedigree objects are
needed for downstream analysis, use `splitped()`.

## 6. Diversity Indicators with `pediv()`

`pediv()` is the integrated diversity summary. It combines:

- founder and ancestor contributions (`fe`, `fa`),
- founder genome equivalents (`fg`),
- three effective population size estimates (`Ne`).

All of these quantities depend on the definition of the **reference
population**. In the present example, the reference population is defined as the
most recent two generations in `deep_ped`.

```{r pediv-setup}
ref_pop <- tp_deep[Gen >= max(Gen) - 1, Ind]
length(ref_pop)
```

```{r pediv-analysis}
div_res <- pediv(tp_deep, reference = ref_pop, top = 10, seed = 42L)

div_res$summary
div_res$ancestors
```

### 6.1 `fe`, `fa`, and `fg`: what do they measure?

These three indicators describe diversity loss from complementary angles.

#### Effective number of founders (`fe`)

$$
f_e = \frac{1}{\sum_{i=1}^{f} p_i^2}
$$

where $p_i$ is the expected contribution of founder $i$ to the reference
population. `fe` answers the question: if all founders had contributed equally,
how many founders would generate the same diversity?

Use `fe` when the main concern is **unequal founder representation**.

#### Effective number of ancestors (`fa`)

$$
f_a = \frac{1}{\sum_{j=1}^{a} q_j^2}
$$

where $q_j$ is the **marginal** contribution of ancestor $j$ after removing the
contributions already explained by more influential ancestors. `fa` is usually
smaller than `fe` when bottlenecks occurred.

Use `fa` when the main concern is **genetic bottlenecks caused by a limited set
of influential ancestors**.

#### Founder genome equivalents (`fg`)

In its simplest interpretation,

$$
f_g = \frac{1}{2\bar{C}}
$$

where $\bar{C}$ is the mean coancestry of the reference population. In
`visPedigree`, `fg` is estimated from a diagonal-corrected mean coancestry:

$$
\hat{\bar{C}} = \frac{N-1}{N}\cdot\frac{\bar{a}_{off}}{2} + \frac{1 + \bar{F}}{2N}
$$

$$
f_g = \frac{1}{2\hat{\bar{C}}}
$$

where $N$ is the size of the full reference population, $\bar{a}_{off}$ is the mean
off-diagonal additive relationship among sampled individuals, and $\bar{F}$
is their mean inbreeding coefficient.

Use `fg` when the main concern is **the amount of founder genome still surviving
after unequal use, bottlenecks, and drift**. In practice, `fg` is often the
most conservative diversity indicator.

### 6.2 Shannon-entropy effective numbers: `feH` and `faH`

The classical `fe` and `fa` are based on $\sum p_i^2$ (Hill number order
$q = 2$), which is disproportionately influenced by a few large contributions.
`visPedigree` also reports the information-theoretic counterpart at order
$q = 1$, derived from Shannon entropy:

$$
f_e^{(H)} = \exp\!\Bigl(-\sum_{i=1}^{f} p_i \ln p_i\Bigr), \qquad
f_a^{(H)} = \exp\!\Bigl(-\sum_{j=1}^{a} q_j \ln q_j\Bigr)
$$

All four metrics satisfy a strict monotone ordering:

$$
N_f \ge f_e^{(H)} \ge f_e, \qquad K \ge f_a^{(H)} \ge f_a
$$

where $N_f$ is the total number of founders and $K$ is the number of ancestors
considered.

#### Metric guide

| Metric | Order | Sensitivity | What it measures |
|---|---|---|---|
| `fe` | $q = 2$ | Dominated by common founders | Effective number of founders if contributions were equalized. Dominated by the largest contributions; many rare founders with small $p_i$ have negligible influence on the metric. |
| `feH` | $q = 1$ | Balanced; sensitive to rare founders | Shannon-entropy effective founders. Counts rare founders more equitably than $q = 2$; always $\ge$ `fe`. |
| `fa` | $q = 2$ | Dominated by common ancestors | Effective number of ancestors accounting for bottleneck structure. Typically lower than `fe`. |
| `faH` | $q = 1$ | Balanced; sensitive to rare ancestors | Shannon-entropy effective ancestors. Less dominated by the single most influential ancestor. |
| `fg` | — | Realized coancestry | Founder genome equivalents; the most conservative indicator because it captures drift as well. |

#### Interpreting the ratio $f_e^{(H)} / f_e$

The ratio $\rho = f_e^{(H)} / f_e$ is a **long-tail signal**:

- $\rho \approx 1$: founder contributions are already concentrated in a small
  number of individuals; there is no meaningful "hidden" diversity from minor
  founders.
- $\rho \gg 1$: many minor founders still carry genetic material into the
  reference population but are effectively invisible to the classical $f_e$
  metric because their contributions are small.  This situation is common early
  in a breeding programme when the pedigree is still shallow.

The analogous ratio $f_a^{(H)} / f_a$ diagnoses the same effect at the
**ancestor** level: a large ratio means that the bottleneck signal from the
dominant ancestor is masking genuine contributions from many secondary
ancestors.

#### Management implications

Use the two ratios together to set priorities:

1. If $f_e^{(H)} / f_e$ is large but $f_e$ itself is already high, the
   programme has good founder coverage; no immediate action needed.
2. If $f_e^{(H)} / f_e$ is large and $f_e$ is low, minor founders carry
   material that could be mobilized — consider deliberately increasing their
   representation in the next generation.
3. If $f_a^{(H)} / f_a$ is large, secondary ancestors are still present but
   marginalized; diversifying sire selection could rebalance the gene pool
   before their lineages are lost entirely.

```{r pediv-shannon}
# Shannon metrics are included in pediv() output
div_res$summary[, .(NFounder, feH, fe, NAncestor, faH, fa)]

# The ratio feH/fe reveals long-tail founder diversity
div_res$summary[, .(rho_founder = feH / fe, rho_ancestor = faH / fa)]
```

```{r pedcontrib-shannon}
# pedcontrib() provides the same metrics via its summary
contrib_res <- pedcontrib(tp_deep, reference = ref_pop, mode = "both")
contrib_res$summary[c("n_founder", "f_e_H", "f_e", "n_ancestor", "f_a_H", "f_a")]
```

## 7. Diversity Dynamics over Time with `pedhalflife()`

`pediv()` provides a static snapshot of diversity. When multiple time points
(generations, years, etc.) are available, `pedhalflife()` tracks how $f_e$,
$f_a$, and $f_g$ evolve and fits a log-linear decay model to quantify the
**rate of diversity loss**.

The total loss rate $\lambda_{total}$ is decomposed into three additive
components that each have a distinct biological meaning:

| Component | Symbol | Source of loss |
|---|---|---|
| Foundation | $\lambda_e$ | Unequal founder contributions |
| Bottleneck | $\lambda_b$ | Overuse of key ancestors |
| Drift | $\lambda_d$ | Random sampling loss |

The **diversity half-life** is the number of time units for $f_g$ to halve:
$$T_{1/2} = \frac{\ln 2}{\lambda_{total}}$$

```{r pedhalflife-analysis}
hl <- pedhalflife(tp_deep, timevar = "Gen")
print(hl)
```

The printed half-life shows the unit in parentheses — `(Gen)` here — taken
directly from the `timevar` argument.  The full object exposes three components:

- `hl$timeseries` — per-time-point table (see column guide below).
- `hl$decay` — single-row table with `LambdaE`, `LambdaB`, `LambdaD`,
  `LambdaTotal`, and `THalf`.
- `hl$timevar` — the `timevar` string passed to the call.

### 7.1 Column guide for `$timeseries`

The cascade rests on a simple algebraic identity:

$$
\ln f_g = \underbrace{\ln f_e}_{\text{LnFe}}
         + \underbrace{\ln\!\left(\frac{f_a}{f_e}\right)}_{\text{LnFaFe}}
         + \underbrace{\ln\!\left(\frac{f_g}{f_a}\right)}_{\text{LnFgFa}}
$$

Because OLS is linear, the slope of $\ln f_g$ versus time equals the sum of the
slopes of the three right-hand terms.  This guarantees exact additivity:
$\lambda_{total} = \lambda_e + \lambda_b + \lambda_d$.

| Column | Formula | What it measures |
|---|---|---|
| `LnFe` | $\ln f_e$ | Log effective-founder diversity. Declines when founder contributions become more unequal over time. |
| `LnFa` | $\ln f_a$ | Log effective-ancestor diversity. Lies below `LnFe` whenever bottleneck ancestors have been over-used. |
| `LnFg` | $\ln f_g$ | Log founder-genome equivalents. The most comprehensive diversity signal; integrates founder use, bottlenecks, and drift. |
| `LnFaFe` | $\ln(f_a/f_e)$ | **Bottleneck gap.** When this ratio decreases over time, a small number of ancestors dominate the gene pool even beyond what unequal founder use would predict. |
| `LnFgFa` | $\ln(f_g/f_a)$ | **Drift gap.** Captures random allele loss due to finite population size. Because $f_g \le f_a$ always, this term is $\le 0$ and becomes more negative as genetic drift accumulates. |
| `TimeStep` | numeric | OLS time axis (same as `Time` for numeric `timevar`; sequential indices otherwise). |

### 7.2 Interpreting the $\lambda$ components

The negative OLS slope of each log column gives the corresponding loss rate.
All three terms have a distinct policy implication:

**$\lambda_e$ (Foundation)** — rate of increase in founder-contribution
inequality.  A large $\lambda_e$ suggests that founder representation is still
diverging, possibly because some founders are reaching later generations
through many more breeding pathways than others.  Remedy: broaden founder
use in the breeding programme.

**$\lambda_b$ (Bottleneck)** — rate at which a few key ancestors capture
an ever-larger share of the gene pool beyond what founder inequality alone
explains.  A large $\lambda_b$ points to systematic over-selection of certain
families or sires.  Remedy: restrict the number of offspring per
high-ranking animal; rotate breeding males.

**$\lambda_d$ (Drift)** — rate of random allele loss in finite populations.
This component is always present and increases with small effective population
size.  It is the only component that cannot be reversed by changing selection
decisions; it can only be slowed by maintaining a larger breeding population
or by cryopreservation of genetic material.

When $\lambda_d \gg \lambda_e + \lambda_b$ (as in the example above), drift
is the dominant driver of diversity loss and should be the primary management
target.

### 7.3 Interpreting $T_{1/2}$

$T_{1/2}$ is the number of time units (generations, years, etc.) required for
$f_g$ to fall to half its current value, assuming the observed decay rate
continues.  It is a single headline figure that combines all three loss
channels.

As a rough benchmark: a half-life shorter than five to ten generations is
generally considered a cause for concern in conservation and aquaculture
genetics.  Breed registries and conservation programmes often use $T_{1/2}$
as a threshold indicator for triggering corrective management.

```{r pedhalflife-timeseries}
hl$timeseries
```

To focus on a specific time window, pass the desired values to the `at`
argument:

```{r pedhalflife-subset}
hl_recent <- pedhalflife(tp_deep, timevar = "Gen",
                         at = tail(sort(unique(tp_deep$Gen)), 4))
print(hl_recent)
```

Plotting the object shows the log-linear decay of each diversity metric:

```{r pedhalflife-plot, fig.width=6, fig.height=4}
plot(hl, type = "log")
```

A `type = "raw"` plot shows the equivalent numbers on the original scale:

```{r pedhalflife-plot-raw, fig.width=6, fig.height=4}
plot(hl, type = "raw")
```

## 8. Effective Population Size with `pedne()` and `pediv()`

`pediv()` reports three complementary effective population size definitions.
Each addresses a distinct biological question.

### 8.1 Demographic `Ne`

$$
N_e = \frac{4 N_m N_f}{N_m + N_f}
$$

where $N_m$ and $N_f$ are the numbers of contributing males and females. This is
the easiest estimate to understand, but it ignores realized pedigree structure,
inbreeding, and drift. It is therefore often optimistic.

### 8.2 Inbreeding `Ne`

$$
\Delta F_i = 1 - (1 - F_i)^{1/(ECG_i - 1)}
$$

$$
N_e = \frac{1}{2\overline{\Delta F}}
$$

This estimator uses the realized rate of inbreeding. `ECG` standardizes
individuals with different pedigree depths. Use this estimate when the primary
concern is **the rate of inbreeding accumulation**.

### 8.3 Coancestry `Ne`

$$
\Delta c_{ij} = 1 - (1 - c_{ij})^{1/(\frac{ECG_i + ECG_j}{2})}
$$

$$
N_e = \frac{1}{2\overline{\Delta c}}
$$

This estimator is based on the rate of coancestry among members of the
reference population. Because it captures the accumulation of relatedness
before it is fully expressed as realized inbreeding, it is often the strictest
and most sensitive warning signal in managed breeding populations.

## 9. Average Relationship Trends with `pedrel()`

`pedrel()` summarizes the mean off-diagonal additive relationship within groups:

$$
MeanRel = \frac{1}{n(n-1)} \sum_{i \ne j} a_{ij}
$$

where $a_{ij} = 2c_{ij}$ is the additive relationship and $c_{ij}$ is the
coancestry (kinship coefficient) between individuals $i$ and $j$.
A higher `MeanRel` means that individuals within the group are, on average,
more related by descent.

This summary is useful for tracking relatedness by generation or year.

```{r pedrel-example}
tp_small$BirthYear <- 2010 + tp_small$Gen

rel_by_gen <- pedrel(tp_small, by = "Gen")
rel_by_gen
```

The `reference` argument lets you focus on a subset of interest inside each
group, such as candidate breeders.

```{r pedrel-reference}
ref_ids <- c("Z1", "Z2", "X", "Y")

pedrel(tp_small, by = "Gen", reference = ref_ids)
```

`MeanRel` and coancestry-based `Ne` are conceptually linked, but they are not
identical summaries. `MeanRel` is an average additive relationship within a
group, whereas coancestry-based `Ne` is derived from the *rate of increase* in
coancestry across pedigree depth.

## 10. Inbreeding Trends with `inbreed()` and `pedfclass()`

Wright's inbreeding coefficient $F$ is the probability that the two alleles of
an individual are identical by descent. A practical starting point is to inspect
the mean trend by generation.

```{r inbreed-trend}
tp_inbred_f <- inbreed(tp_inbred)

f_trend <- tp_inbred_f[, .(MeanF = mean(f, na.rm = TRUE)), by = Gen]
f_trend
```

For classification by inbreeding severity, `pedfclass()` can be applied
directly to the pedigree object. If the inbreeding coefficient is not yet
present, the function computes it internally.

```{r pedfclass-example}
pedfclass(tp_inbred)
```

Custom reporting thresholds can be supplied through `breaks`.

```{r pedfclass-breaks}
pedfclass(tp_inbred, breaks = c(0.03125, 0.0625, 0.125, 0.25))
```

The default thresholds correspond approximately to familiar pedigree scenarios:

- $F = 0.0625$: half-sib mating.
- $F = 0.125$: avuncular or grandparent-grandchild mating.
- $F = 0.25$: full-sib or parent-offspring mating.

## 11. Gene Flow and Partial Inbreeding

### 11.1 `pedancestry()`: founder-line proportions

`pedancestry()` tracks how founder groups contribute to later descendants. This
is useful when founders are labeled by line, strain, or geographic origin.

```{r pedancestry-example}
ped_line <- data.table(
  Ind = c("A", "B", "C", "D", "E", "F", "G"),
  Sire = c(NA, NA, NA, NA, "A", "C", "E"),
  Dam  = c(NA, NA, NA, NA, "B", "D", "F"),
  Sex = c("male", "female", "male", "female", "male", "female", "male"),
  Line = c("Line1", "Line1", "Line2", "Line2", NA, NA, NA)
)

tp_line <- tidyped(ped_line)

anc <- pedancestry(tp_line, foundervar = "Line")
anc
```

### 11.2 `pedpartial()`: which ancestors explain inbreeding?

`pedpartial()` decomposes total inbreeding into contributions from selected
ancestors. It is useful for identifying which ancestors are most responsible
for the observed inbreeding.

```{r pedpartial-example}
partial_f <- pedpartial(tp_inbred, ancestors = c("A", "B"))
partial_f
```

## 12. Practical Interpretation

One useful interpretation order is:

1. Check pedigree depth and completeness with `pedecg()`.
2. Check generation timing with `pedgenint()`.
3. Quantify static diversity with `fe`, `fa`, and `fg` via `pediv()`; compare
   with `feH` and `faH` to assess long-tail founder value.
4. Track diversity dynamics over time with `pedhalflife()`. Inspect
   $\lambda_e$, $\lambda_b$, and $\lambda_d$ to identify the dominant driver
   of loss; use $T_{1/2}$ as a headline indicator for management reporting.
5. Compare demographic, inbreeding, and coancestry-based `Ne`.
6. Monitor `MeanRel` and `MeanF` over time.
7. Use `pedsubpop()`, `pedancestry()`, and `pedpartial()` to diagnose structure,
   introgression, and bottlenecks.

## References

- Boichard, D., Maignel, L., & Verrier, É. (1997). The value of using probabilities of gene origin to measure genetic variability in a population. *Genetics Selection Evolution*, 29(1), 5-23.
- Caballero, A., & Toro, M. A. (2000). Interrelations between effective population size and other pedigree tools for the management of conserved populations. *Genetical Research*, 75(3), 331-343.
- Cervantes, I., Goyache, F., Molina, A., Valera, M., & Gutiérrez, J. P. (2011). Estimation of effective population size from the rate of coancestry in pedigreed populations. *Journal of Animal Breeding and Genetics*, 128(1), 56-63.
- Gutiérrez, J. P., Cervantes, I., Molina, A., Valera, M., & Goyache, F. (2008). Individual increase in inbreeding allows estimating effective sizes from pedigrees. *Genetics Selection Evolution*, 40(4), 359-378.
- Gutiérrez, J. P., Cervantes, I., & Goyache, F. (2009). Improving the estimation of realized effective population sizes in farm animals. *Journal of Animal Breeding and Genetics*, 126(4), 327-332.
- Hill, M. O. (1973). Diversity and evenness: a unifying notation and its consequences. *Ecology*, 54(2), 427-432.
- Jost, L. (2006). Entropy and diversity. *Oikos*, 113(2), 363-375.
- Lacy, R. C. (1989). Analysis of founder representation in pedigrees: founder equivalents and founder genome equivalents. *Zoo Biology*, 8(2), 111-123.
- Wright, S. (1922). Coefficients of inbreeding and relationship. *The American Naturalist*, 56(645), 330-338.
- Wright, S. (1931). Evolution in Mendelian populations. *Genetics*, 16(2), 97-159.