---
title: "Pedigree Analysis and Population Genetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pedigree Analysis and Population Genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette summarizes a practical workflow for pedigree analysis with
`visPedigree`, with emphasis on the interpretation of the main indicators used
in breeding and conservation genetics.

The discussion is organized around five questions:

1. How complete and deep is the pedigree?
2. How long is the generation cycle?
3. Has genetic diversity been eroded by unequal founder use, bottlenecks, or drift?
4. How large is the effective population size under different definitions?
5. Are relationship, inbreeding, subpopulation structure, and gene flow under control?

## 1. Setup and Data Preparation

Different package datasets are used for different analytical tasks.

- `deep_ped` is useful for pedigree depth and diversity summaries.
- `big_family_size_ped` contains a `Year` column and is suitable for generation intervals.
- `small_ped` is convenient for relationship examples.
- `inbred_ped` is convenient for inbreeding and classification examples.

```{r setup}
library(visPedigree)
library(data.table)

data(deep_ped, package = "visPedigree")
data(big_family_size_ped, package = "visPedigree")
data(small_ped, package = "visPedigree")
data(inbred_ped, package = "visPedigree")

tp_deep <- tidyped(deep_ped)
tp_small <- tidyped(small_ped)
tp_inbred <- tidyped(inbred_ped)
```

## 2. Pedigree Overview with `pedstats()`

`pedstats()` provides a compact structural summary with three components:

- `$summary`: pedigree size and parental structure.
- `$ecg`: pedigree completeness and ancestral depth.
- `$gen_intervals`: generation intervals, only when a usable time column is available.

Here `deep_ped` has no explicit birth-date column. Accordingly,
`pedstats()` returns `$summary` and `$ecg`, whereas `$gen_intervals` remains
`NULL`.

```{r pedstats-overview}
stats_deep <- pedstats(tp_deep)

stats_deep$summary
tail(stats_deep$ecg)
stats_deep$gen_intervals
```

Pedigree depth and pedigree time are related but distinct quantities. The former
is addressed by `pedecg()`, whereas the latter is addressed by `pedgenint()`.

## 3. Pedigree Completeness with `pedecg()`

Equivalent complete generations (ECG) summarize the amount of ancestral
information available for each individual:

$$
ECG_i = \sum_j \left(\frac{1}{2}\right)^{g_{ij}}
$$

where $g_{ij}$ is the number of generations between individual $i$ and ancestor
$j$. ECG increases with both pedigree depth and pedigree completeness.

In practice:

- `ECG` combines depth and completeness.
- `FullGen` counts how many fully known ancestral generations exist.
- `MaxGen` records the deepest known ancestral path.

`ECG` is especially useful because it provides the depth adjustment required by
realized effective population size estimators based on inbreeding or coancestry.

```{r pedecg-example}
ecg_deep <- pedecg(tp_deep)

ecg_deep[order(-ECG)][1:10]
```

## 4. Generation Intervals with `pedgenint()`

Generation interval is the age of a parent at the birth of its offspring:

$$
L = t_{offspring} - t_{parent}
$$

`pedgenint()` estimates this quantity for seven pathway summaries:

- `SS`, `SD`, `DS`, `DD`: sex-specific gametic pathways.
- `SO`, `DO`: sex-independent sire-offspring and dam-offspring pathways.
- `Average`: all parent-offspring pairs combined.

The function accepts `Date`/`POSIXct` columns, date strings, or numeric years.
When only an integer year is available, it is converted internally to
`YYYY-07-01` as a mid-year approximation.

```{r pedgenint-basic}
tp_time <- tidyped(big_family_size_ped)

genint_year <- pedgenint(tp_time, timevar = "Year", unit = "year")
genint_year
```

The optional `cycle` parameter adds `GenEquiv`, which compares the observed mean
interval with a target breeding cycle:

$$
GenEquiv_i = \frac{\bar{L}_i}{L_{cycle}}
$$

Values larger than 1 indicate that the realized generation interval exceeds the
target cycle.

```{r pedgenint-cycle}
genint_cycle <- pedgenint(tp_time, timevar = "Year", unit = "year", cycle = 1.2)

genint_cycle[Pathway %in% c("SS", "SD", "DS", "DD", "Average")]
```

## 5. Subpopulation Structure with `pedsubpop()`

Before interpreting diversity or relationship metrics, it is useful to check
whether the pedigree forms a single connected population or a mixture of
separate components.

`pedsubpop()` has two modes:

- `by = NULL`: summarize disconnected pedigree components.
- `by = "..."`: summarize the pedigree by an existing grouping variable.

```{r pedsubpop-example}
ped_demo <- data.table(
  Ind = c("A", "B", "C", "D", "E", "F", "G", "H"),
  Sire = c(NA, NA, "A", NA, NA, "E", NA, NA),
  Dam  = c(NA, NA, "B", NA, NA, NA, NA, NA),
  Sex = c("male", "female", "male", "female", "male", "female", "male", "female"),
  Batch = c("L1", "L1", "L1", "L1", "L2", "L2", "L3", "L3")
)

tp_demo <- tidyped(ped_demo)

pedsubpop(tp_demo)
pedsubpop(tp_demo, by = "Batch")
```

This is a compact summary tool. When the actual split pedigree objects are
needed for downstream analysis, use `splitped()`.

## 6. Diversity Indicators with `pediv()`

`pediv()` is the integrated diversity summary. It combines:

- founder and ancestor contributions (`fe`, `fa`),
- founder genome equivalents (`fg`),
- three effective population size estimates (`Ne`).

All of these quantities depend on the definition of the **reference
population**. In the present example, the reference population is defined as the
most recent two generations in `deep_ped`.

```{r pediv-setup}
ref_pop <- tp_deep[Gen >= max(Gen) - 1, Ind]
length(ref_pop)
```

```{r pediv-analysis}
div_res <- pediv(tp_deep, reference = ref_pop, top = 10, seed = 42L)

div_res$summary
div_res$ancestors
```

### 6.1 `fe`, `fa`, and `fg`: what do they measure?

These three indicators describe diversity loss from complementary angles.

#### Effective number of founders (`fe`)

$$
f_e = \frac{1}{\sum_{i=1}^{f} p_i^2}
$$

where $p_i$ is the expected contribution of founder $i$ to the reference
population. `fe` answers the question: if all founders had contributed equally,
how many founders would generate the same diversity?

Use `fe` when the main concern is **unequal founder representation**.

#### Effective number of ancestors (`fa`)

$$
f_a = \frac{1}{\sum_{j=1}^{a} q_j^2}
$$

where $q_j$ is the **marginal** contribution of ancestor $j$ after removing the
contributions already explained by more influential ancestors. `fa` is usually
smaller than `fe` when bottlenecks occurred.

Use `fa` when the main concern is **genetic bottlenecks caused by a limited set
of influential ancestors**.

#### Founder genome equivalents (`fg`)

In its simplest interpretation,

$$
f_g = \frac{1}{2\bar{C}}
$$

where $\bar{C}$ is the mean coancestry of the reference population. In
`visPedigree`, `fg` is estimated from a diagonal-corrected mean coancestry:

$$
\hat{\bar{C}} = \frac{N-1}{N}\cdot\frac{\bar{a}_{off}}{2} + \frac{1 + \bar{F}_s}{2N}
$$

$$
f_g = \frac{1}{2\hat{\bar{C}}}
$$

where $N$ is the size of the full reference cohort, $\bar{a}_{off}$ is the mean
off-diagonal additive relationship among sampled individuals, and $\bar{F}_s$
is their mean inbreeding coefficient.

Use `fg` when the main concern is **the amount of founder genome still surviving
after unequal use, bottlenecks, and drift**. In practice, `fg` is often the
most conservative diversity indicator.

## 7. Effective Population Size with `pedne()` and `pediv()`

`pediv()` reports three complementary effective population size definitions.
Each addresses a distinct biological question.

### 7.1 Demographic `Ne`

$$
N_e = \frac{4 N_m N_f}{N_m + N_f}
$$

where $N_m$ and $N_f$ are the numbers of contributing males and females. This is
the easiest estimate to understand, but it ignores realized pedigree structure,
inbreeding, and drift. It is therefore often optimistic.

### 7.2 Inbreeding `Ne`

$$
\Delta F_i = 1 - (1 - F_i)^{1/(ECG_i - 1)}
$$

$$
N_e = \frac{1}{2\overline{\Delta F}}
$$

This estimator uses the realized rate of inbreeding. `ECG` standardizes
individuals with different pedigree depths. Use this estimate when the primary
concern is **the rate of inbreeding accumulation**.

### 7.3 Coancestry `Ne`

$$
\Delta c_{ij} = 1 - (1 - c_{ij})^{1/(\frac{ECG_i + ECG_j}{2})}
$$

$$
N_e = \frac{1}{2\overline{\Delta c}}
$$

This estimator is based on the rate of coancestry among members of the
reference population. Because it captures the accumulation of relatedness
before it is fully expressed as realized inbreeding, it is often the strictest
and most sensitive warning signal in managed breeding populations.

## 8. Average Relationship Trends with `pedrel()`

`pedrel()` summarizes the mean off-diagonal additive relationship within groups:

$$
MeanRel = \frac{1}{n(n-1)} \sum_{i \ne j} a_{ij}
$$

where $a_{ij} = 2f_{ij}$. A higher `MeanRel` means that individuals within the
group are, on average, more related by descent.

This summary is useful for tracking relatedness by generation or year.

```{r pedrel-example}
tp_small$BirthYear <- 2010 + tp_small$Gen

rel_by_gen <- pedrel(tp_small, by = "Gen")
rel_by_gen
```

The `reference` argument lets you focus on a subset of interest inside each
group, such as candidate breeders.

```{r pedrel-reference}
ref_ids <- c("Z1", "Z2", "X", "Y")

pedrel(tp_small, by = "Gen", reference = ref_ids)
```

`MeanRel` and coancestry-based `Ne` are conceptually linked, but they are not
identical summaries. `MeanRel` is an average additive relationship within a
group, whereas coancestry-based `Ne` is derived from the *rate of increase* in
coancestry across pedigree depth.

## 9. Inbreeding Trends with `inbreed()` and `pedfclass()`

Wright's inbreeding coefficient $F$ is the probability that the two alleles of
an individual are identical by descent. A practical starting point is to inspect
the mean trend by generation.

```{r inbreed-trend}
tp_inbred_f <- inbreed(tp_inbred)

f_trend <- tp_inbred_f[, .(MeanF = mean(f, na.rm = TRUE)), by = Gen]
f_trend
```

For classification by inbreeding severity, `pedfclass()` can be applied
directly to the pedigree object. If the inbreeding coefficient is not yet
present, the function computes it internally.

```{r pedfclass-example}
pedfclass(tp_inbred)
```

Custom reporting thresholds can be supplied through `breaks`.

```{r pedfclass-breaks}
pedfclass(tp_inbred, breaks = c(0.03125, 0.0625, 0.125, 0.25))
```

The default thresholds correspond approximately to familiar pedigree scenarios:

- $F = 0.0625$: half-sib mating.
- $F = 0.125$: avuncular or grandparent-grandchild mating.
- $F = 0.25$: full-sib or parent-offspring mating.

## 10. Gene Flow and Partial Inbreeding

### 10.1 `pedancestry()`: founder-line proportions

`pedancestry()` tracks how founder groups contribute to later descendants. This
is useful when founders are labeled by line, strain, or geographic origin.

```{r pedancestry-example}
ped_line <- data.table(
  Ind = c("A", "B", "C", "D", "E", "F", "G"),
  Sire = c(NA, NA, NA, NA, "A", "C", "E"),
  Dam  = c(NA, NA, NA, NA, "B", "D", "F"),
  Sex = c("male", "female", "male", "female", "male", "female", "male"),
  Line = c("Line1", "Line1", "Line2", "Line2", NA, NA, NA)
)

tp_line <- tidyped(ped_line)

anc <- pedancestry(tp_line, foundervar = "Line")
anc
```

### 10.2 `pedpartial()`: which ancestors explain inbreeding?

`pedpartial()` decomposes total inbreeding into contributions from selected
ancestors. It is useful for identifying which ancestors are most responsible
for the observed inbreeding.

```{r pedpartial-example}
partial_f <- pedpartial(tp_inbred, ancestors = c("A", "B"))
partial_f
```

## 11. Practical Interpretation

One useful interpretation order is:

1. Check pedigree depth and completeness with `pedecg()`.
2. Check generation timing with `pedgenint()`.
3. Quantify diversity loss with `fe`, `fa`, and `fg`.
4. Compare demographic, inbreeding, and coancestry-based `Ne`.
5. Monitor `MeanRel` and `MeanF` over time.
6. Use `pedsubpop()`, `pedancestry()`, and `pedpartial()` to diagnose structure,
   introgression, and bottlenecks.

## References

- Boichard, D., Maignel, L., & Verrier, É. (1997). The value of using probabilities of gene origin to measure genetic variability in a population. *Genetics Selection Evolution*, 29(1), 5-23.
- Caballero, A., & Toro, M. A. (2000). Interrelations between effective population size and other pedigree tools for the management of conserved populations. *Genetical Research*, 75(3), 331-343.
- Cervantes, I., Goyache, F., Molina, A., Valera, M., & Gutiérrez, J. P. (2011). Estimation of effective population size from the rate of coancestry in pedigreed populations. *Journal of Animal Breeding and Genetics*, 128(1), 56-63.
- Gutiérrez, J. P., Cervantes, I., Molina, A., Valera, M., & Goyache, F. (2008). Individual increase in inbreeding allows estimating effective sizes from pedigrees. *Genetics Selection Evolution*, 40(4), 359-370.
- Gutiérrez, J. P., Cervantes, I., & Goyache, F. (2009). Improving the estimation of realized effective population sizes in farm animals. *Journal of Animal Breeding and Genetics*, 126(4), 327-332.
- Lacy, R. C. (1989). Analysis of founder representation in pedigrees: founder equivalents and founder genome equivalents. *Zoo Biology*, 8(2), 111-123.
- Wright, S. (1922). Coefficients of inbreeding and relationship. *The American Naturalist*, 56(645), 330-338.
- Wright, S. (1931). Evolution in Mendelian populations. *Genetics*, 16(2), 97-159.