---
title: "5. Calculation and visualization of relationship matrix"
author: "Sheng Luan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Calculation and visualization of relationship matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 6.5,
  dpi = 300,
  out.width = "100%"
)
library(visPedigree)
library(Matrix)
```

1. [Calculating Relationship Matrices with pedmat()](#1)   
1.1 [Supported Methods](#1-1)   
1.2 [Basic Usage](#1-2)   
1.3 [Sparse Matrix Representation](#1-3)   
2. [Inspecting the Matrix](#2)   
2.1 [Summary Statistics](#2-1)   
2.2 [Querying Specific Relationships](#2-2)   
3. [Compact Mode for Large Pedigrees](#3)   
3.1 [Using compact = TRUE](#3-1)   
3.2 [Expanding and Querying Compacted Matrices](#3-2)   
3.3 [When to Use Compact Mode](#3-3)   
4. [Visualizing Relationship Matrices with vismat()](#4)   
4.1 [Relationship Heatmaps](#4-1)   
4.2 [Inbreeding and Kinship Histograms](#4-2)   
5. [Performance Considerations](#5)

Relationship matrices are fundamental tools in quantitative genetics and animal breeding. They quantify the genetic similarity between individuals due to shared ancestry, which is essential for estimating breeding values (BLUP) and managing genetic diversity. The `visPedigree` package provides efficient tools for calculating various relationship matrices and visualizing them through heatmaps and histograms.

## 1. Calculating Relationship Matrices with `pedmat()` {#1}

The `pedmat()` function is the primary tool for calculating relationship matrices. It supports both additive and dominance relationship matrices, as well as their inverses.

### 1.1 Supported Methods {#1-1}

The `method` parameter in `pedmat()` determines the type of matrix to calculate:

-   **"A"**: Additive relationship matrix (Numerator Relationship Matrix).
-   **"Ainv"**: Inverse of the additive relationship matrix.
-   **"D"**: Dominance relationship matrix.
-   **"Dinv"**: Inverse of the dominance relationship matrix.
-   **"AA"**: Additive-by-additive (epistatic) relationship matrix.
-   **"AAinv"**: Inverse of the epistatic relationship matrix.
-   **"f"**: Inbreeding coefficients vector (uses the same optimized engine as `tidyped(..., inbreed = TRUE)`).

### 1.2 Basic Usage {#1-2}

Most calculations require a pedigree tidied by `tidyped()`.

```{r basic_calc}
# Load example pedigree and tidy it
data(small_ped)
tped <- tidyped(small_ped)

# Calculate Additive Relationship Matrix (A)
mat_A <- pedmat(tped, method = "A")

# Calculate Dominance Relationship Matrix (D)
mat_D <- pedmat(tped, method = "D")

# Calculate inbreeding coefficients (f)
vec_f <- pedmat(tped, method = "f")
```

### 1.3 Sparse Matrix Representation {#1-3}

By default, `pedmat()` returns a sparse matrix (class `dsCMatrix` from the `Matrix` package) for relationship matrices. This is highly memory-efficient for large pedigrees where many individuals are unrelated.

```{r sparse_check}
class(mat_A)
```

## 2. Inspecting the Matrix {#2}

### 2.1 Summary Statistics {#2-1}

Use the `summary()` method to get an overview of the calculated matrix, including size, density, and average relationship.

```{r matrix_summary}
tail(summary(mat_A),10)
```

### 2.2 Querying Specific Relationships {#2-2}

Instead of manually indexing the matrix, you can use `query_relationship()` to retrieve coefficients by individual IDs.

```{r query}
# Query relationship between Z1 and Z2
query_relationship(mat_A, "Z1", "Z2")

# Query multiple pairs
query_relationship(mat_A, c("Z1", "A"), c("Z2", "B"))
```

## 3. Compact Mode for Large Pedigrees {#3}

For large pedigrees with many full-sibling families (common in aquatic breeding populations), `pedmat()` can merge full siblings into representative nodes to save memory and time.

### 3.1 Using `compact = TRUE` {#3-1}

When `compact = TRUE`, the matrix is calculated for unique representative individuals from each full-sib family.

```{r compact_calc}
# Calculate compacted A matrix
mat_compact <- pedmat(tped, method = "A", compact = TRUE)

# The result is a 'pedmat' object containing the compacted matrix
print(mat_compact[10,10])
```

### 3.2 Expanding and Querying Compacted Matrices {#3-2}

If you need the full matrix after a compact calculation, use `expand_pedmat()`. For retrieving specific values, `query_relationship()` handles both standard and compact objects transparently.

```{r expand}
# Expand to full 28x28 matrix
mat_full <- expand_pedmat(mat_compact)
dim(mat_full)

# Query still works the same way
query_relationship(mat_compact, "Z1", "Z2")
```

### 3.3 When to Use Compact Mode {#3-3}

Compact mode is highly recommended for:

*   **Large Pedigrees**: More than 5,000 individuals with substantial full-sibling groups.
*   **High-fecundity species**: Such as aquatic animals or plants, where families often have hundreds or thousands of offspring.
*   **Memory-limited environments**: When the full matrix exceeds available RAM.

| Pedigree Size | Full-Sib Proportion | Recommended Mode |
| :--- | :--- | :--- |
| < 1,000 | Any | Standard |
| > 5,000 | < 20% | Standard / Compact |
| > 5,000 | > 20% | **Compact** |

## 4. Visualizing Relationship Matrices with `vismat()` {#4}

Visualization helps in understanding population structure, detecting family clusters, and checking the distribution of genetic relationships.

### 4.1 Relationship Heatmaps {#4-1}

The `"heatmap"` type (default) uses a Nature Genetics style color palette
(White–Orange–Red) to display relationships. Rows and columns are reordered
by hierarchical clustering (Ward.D2) by default, bringing closely related
individuals into contiguous blocks — full-sibs cluster tightly because they
share nearly identical relationship profiles with the rest of the population.

```{r heatmap, fig.width=6, fig.height=6}
# Heatmap of the A matrix (with default clustering reorder)
vismat(mat_A)
```

#### Compact Matrix — Direct Visualization

A compact `pedmat` object can be passed directly to `vismat()`. It is
automatically expanded to full dimensions before rendering.

```{r heatmap_compact, fig.width=6, fig.height=6}
# Compact matrix: expanded automatically (message printed)
vismat(mat_compact)
```

#### Preserve Pedigree Order

Set `reorder = FALSE` to keep the original pedigree order instead of
re-sorting by clustering.

```{r heatmap_no_reorder, fig.width=6, fig.height=6}
vismat(mat_A, reorder = FALSE)
```

#### Display a Subset of Individuals

Use `ids` to focus on specific individuals.

```{r heatmap_ids, fig.width=5, fig.height=5}
target_ids <- rownames(as.matrix(mat_A))[1:8]
vismat(mat_A, ids = target_ids,
       main = "Relationship Heatmap — First 8 Individuals")
```

#### Grouping by Pedigree Column

For large populations, aggregate relationships to a group-level view using
the `by` parameter. The matrix is reduced to mean coefficients between groups.

```{r heatmap_group, fig.width=6, fig.height=6}
# Mean relationship between generations
vismat(mat_A, ped = tped, by = "Gen",
       main = "Mean Relationship Between Generations")
```

```{r heatmap_family, fig.width=6, fig.height=6}
# Mean relationship between full-sib families
# (founders without a family assignment are excluded automatically)
vismat(mat_A, ped = tped, by = "Family",
       main = "Mean Relationship Between Full-Sib Families")
```

### 4.2 Inbreeding and Kinship Histograms {#4-2}

The "histogram" type displays the distribution of relationship coefficients (lower triangle) or inbreeding coefficients.

```{r histogram, fig.width=6, fig.height=4}
# Distribution of relationship coefficients
vismat(mat_A, type = "histogram")
```

## 5. Performance Considerations {#5}

Calculation and visualization of large matrices can be resource-intensive.
`vismat()` applies the following automatic optimizations:

| Condition | Behavior |
| :--- | :--- |
| Compact + `by` | Group means are computed directly from the compact matrix (no full expansion) |
| Compact, no `by`, N > 5 000 | Uses compact representative view (labels show `ID (×n)`) |
| Compact, no `by`, N ≤ 5 000 | Matrix is automatically expanded via `expand_pedmat()` |
| N > 2 000 | Hierarchical clustering (reorder) is automatically skipped |
| N > 500 | Individual labels are automatically hidden |
| N > 100 | Grid lines are automatically hidden |

When a compact `pedmat` is used with `by`, `vismat()` computes the group-level
mean relationship matrix algebraically from the K×K compact matrix, including
a sibling off-diagonal correction. This avoids expanding to the full N×N matrix,
making family-level or generation-level visualization feasible even for pedigrees
with hundreds of thousands of individuals.

The example below uses `big_family_size_ped` (178 431 individuals, compact to 2 626)
and displays the mean additive relationship among **all** full-sib families in the
latest generation — a computation that would be infeasible with full expansion.

```{r large_ped_tip, fig.width=7, fig.height=6.5}
data(big_family_size_ped)

tp_big <- tidyped(big_family_size_ped)
last_gen <- max(tp_big$Gen, na.rm = TRUE)

# Compute the compact A matrix for the entire pedigree
mat_big_compact <- pedmat(tp_big, method = "A", compact = TRUE)

# Focus on all individuals in the last generation that belong to a family
ids_last_gen <- tp_big[Gen == last_gen & !is.na(Family), Ind]

# vismat() aggregates directly from the compact matrix — no expansion needed
vismat(
       mat_big_compact,
       ped = tp_big,
       ids = ids_last_gen,
       by = "Family",
       main = paste("Mean Relationship Between All Families in Generation", last_gen)
)
```

This family-level view reveals the genetic structure among all `r length(unique(tp_big[Gen == last_gen & !is.na(Family), Family]))` families
comprising `r length(ids_last_gen)` individuals, computed in seconds from the
compact matrix.

---
**See Also:**
- `vignette("tidy-pedigree", package = "visPedigree")`
- `vignette("draw-pedigree", package = "visPedigree")`
